---
title: "Liver proteomics analysis using DEP2"
author: "Ellen Mazalale"
date: "06022026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
## Load packages and define analysis-wide settings.

suppressPackageStartupMessages({
  library(tidyverse)
  library(SummarizedExperiment)
  library(DEP2)
  library(limma)
  library(vsn)
  library(ggplot2)
  library(pheatmap)
  library(grid)
})

pg_file <- "merged_pg_matrix.tsv"
meta_file <- "merged_sample_metadata.csv"

OUT_DIR <- "liver_DEP2_outputs"
dir.create(OUT_DIR, showWarnings = FALSE)

MIN_DETECT_FRAC <- 0.50
FDR_CUTOFF <- 0.05

set.seed(1)
```

# Analysis objective

This analysis subsets the merged proteomics data to liver samples only and uses DEP2 for protein identifier handling, SummarizedExperiment construction, missing-value filtering, VSN normalization, imputation, and quality control. The downstream modeling is performed on the DEP2-processed abundance matrix using limma so that age, environment, sex, and the age-by-environment interaction can be evaluated directly. The main biological goal is to determine whether liver aging from 8 to 18 weeks is detectable in laboratory and mesocosm environments, whether the direction or magnitude of aging differs between environments, and whether any dominant biological pattern, such as ribosomal remodeling, emerges from the age-associated proteins.

# 1. Load raw matrix and metadata

```{r load-data}
## Load the merged protein-group matrix and sample metadata.

mat_raw <- readr::read_tsv(pg_file, show_col_types = FALSE)
meta_raw <- readr::read_csv(meta_file, show_col_types = FALSE)

cat("Raw protein matrix dimensions:", dim(mat_raw), "\n")
cat("Raw metadata dimensions:", dim(meta_raw), "\n")

head(meta_raw)
```

# 2. Subset to liver samples

```{r subset-liver}
## Restrict metadata to liver samples from the 8-week and 18-week time points.

meta_liver <- meta_raw %>%
  dplyr::filter(tissue == "Liver") %>%
  dplyr::filter(age %in% c("8wk", "18wk")) %>%
  dplyr::mutate(
    age = factor(age, levels = c("8wk", "18wk")),
    environment = factor(environment, levels = c("Lab", "Mesocosm")),
    sex = factor(sex, levels = c("Female", "Male"))
  )

cat("liver samples retained:", nrow(meta_liver), "\n")

table(meta_liver$age, meta_liver$environment, meta_liver$sex)
```


```{r}
## Confirm that every liver sample column listed in metadata is present in the protein matrix.

missing_sample_cols <- setdiff(meta_liver$sample_column, colnames(mat_raw))

cat("Missing sample columns:", length(missing_sample_cols), "\n")

if (length(missing_sample_cols) > 0) {
  print(missing_sample_cols)
}
```


```{r}
## Create the experimental design table for DEP2 using liver metadata.

exp_design_liver <- meta_liver %>%
  dplyr::arrange(group_name, sample) %>%
  dplyr::group_by(group_name) %>%
  dplyr::mutate(
    condition = as.character(group_name),
    replicate = dplyr::row_number()
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    label = sample_column,
    condition,
    replicate,
    tissue,
    age,
    environment,
    sex,
    sample,
    group_name
  )

print(exp_design_liver, n = Inf)
```


```{r}
## Create the liver-specific protein matrix.

sample_cols_liver <- exp_design_liver$label

mat_liver <- mat_raw %>%
  dplyr::select(
    Protein.Group,
    Protein.Names,
    Genes,
    First.Protein.Description,
    N.Sequences,
    N.Proteotypic.Sequences,
    dplyr::all_of(sample_cols_liver)
  )

cat("liver matrix dimensions:", dim(mat_liver), "\n")

cat(
  "Sample columns match experimental design:",
  setequal(
    sample_cols_liver,
    colnames(mat_liver)[-(1:6)]
  ),
  "\n"
)
```


```{r}
## Remove protein groups with no quantified values in any liver sample.

keep_liver_proteins <- rowSums(
  !is.na(mat_liver[, sample_cols_liver])
) > 0

cat("Proteins before removing all-missing liver rows:", nrow(mat_liver), "\n")
cat("Proteins detected in at least one liver sample:", sum(keep_liver_proteins), "\n")
cat("Proteins absent from all liver samples:", sum(!keep_liver_proteins), "\n")

mat_liver <- mat_liver[keep_liver_proteins, ]

cat("liver matrix after removing all-missing rows:", dim(mat_liver), "\n")
```


```{r}
## Prepare protein identifiers for DEP2 by creating clean gene names and unique protein IDs.

mat_dep <- mat_liver %>%
  dplyr::mutate(
    Genes_clean = ifelse(
      is.na(Genes) | Genes == "",
      Protein.Group,
      Genes
    )
  )

data_unique <- make_unique(
  mat_dep,
  names = "Genes_clean",
  ids = "Protein.Group",
  delim = ";"
)

cat("Input liver proteins:", nrow(mat_dep), "\n")
cat("Unique proteins after make_unique:", nrow(data_unique), "\n")
cat("Missing names after cleanup:", sum(is.na(data_unique$name) | data_unique$name == ""), "\n")
cat("Missing IDs after cleanup:", sum(is.na(data_unique$ID) | data_unique$ID == ""), "\n")
```

##Just checking
```{r}
dim(data_se)
dim(data_filt)
ncol(data_se)

```



```{r}
## Create a DEP2 SummarizedExperiment object from the liver-specific protein matrix.

data_se <- make_se(
  data_unique,
  columns = exp_design_liver$label,
  expdesign = exp_design_liver
)

cat("Proteins in DEP2 object:", nrow(data_se), "\n")
cat("Samples in DEP2 object:", ncol(data_se), "\n")
```


```{r}
## Summarize protein completeness before DEP2 filtering.

det_counts <- rowSums(!is.na(assay(data_se)))

summary(det_counts)

cat(
  "Proteins detected in all 38 samples:",
  sum(det_counts == 38),
  "\n"
)

cat(
  "Proteins detected in at least 50% of samples:",
  sum(det_counts >= 19),
  "\n"
)

cat(
  "Proteins detected in only one sample:",
  sum(det_counts == 1),
  "\n"
)
```


```{r}
## Filter proteins based on replicate completeness within conditions.

data_filt <- filter_se(
  data_se,
  thr = MIN_DETECT_FRAC
)

cat("Proteins before filtering:", nrow(data_se), "\n")
cat("Proteins after filtering:", nrow(data_filt), "\n")
cat("Proteins removed:", nrow(data_se) - nrow(data_filt), "\n")
```


```{r}
## Examine how many proteins were retained by DEP filtering.

cat(
  "Percent retained:",
  round(100 * nrow(data_filt) / nrow(data_se), 1),
  "%\n"
)

cat(
  "Percent removed:",
  round(100 * (nrow(data_se) - nrow(data_filt)) / nrow(data_se), 1),
  "%\n"
)
```


```{r}
## Apply variance stabilizing normalization (VSN).

data_norm <- normalize_vsn(data_filt)

cat("Proteins after normalization:", nrow(data_norm), "\n")
cat("Samples after normalization:", ncol(data_norm), "\n")
```


```{r}
## Visualize sample intensity distributions after VSN normalization.

plot_normalization(data_norm)
```



```{r}
## Quantify missing values before imputation.

miss_by_sample <- data.frame(
  sample = colnames(assay(data_norm)),
  frac_missing = colMeans(is.na(assay(data_norm)))
)

summary(miss_by_sample$frac_missing)

cat(
  "Mean sample missingness:",
  round(mean(miss_by_sample$frac_missing) * 100, 1),
  "%\n"
)

cat(
  "Maximum sample missingness:",
  round(max(miss_by_sample$frac_missing) * 100, 1),
  "%\n"
)
```


```{r}
## Identify samples with the highest missingness.

miss_by_sample %>%
  dplyr::arrange(desc(frac_missing)) %>%
  head(10)
```




```{r}
## Create a metadata table using the DEP2 sample names.

sample_meta <- as.data.frame(colData(data_norm))
sample_meta$sample <- rownames(sample_meta)

head(sample_meta[, c("sample", "condition", "sex", "age", "environment")])
```


```{r}
## Compare missingness by sex using DEP2 sample names.

miss_by_sample <- data.frame(
  sample = colnames(assay(data_norm)),
  frac_missing = colMeans(is.na(assay(data_norm)))
)

sample_meta <- as.data.frame(colData(data_norm))
sample_meta$sample <- rownames(sample_meta)

miss_by_sample <- miss_by_sample %>%
  dplyr::left_join(
    sample_meta %>%
      dplyr::select(sample, condition, sex, age, environment),
    by = "sample"
  )

miss_by_sample %>%
  dplyr::group_by(sex) %>%
  dplyr::summarise(
    mean_missing = mean(frac_missing),
    sd_missing = sd(frac_missing),
    n = dplyr::n()
  )
```


```{r}
## Visualize missingness by sex and environment.

ggplot(
  miss_by_sample,
  aes(x = sex, y = frac_missing, fill = environment)
) +
  geom_boxplot(outlier.alpha = 0.4) +
  geom_jitter(
    width = 0.15,
    size = 2,
    alpha = 0.8
  ) +
  theme_bw() +
  labs(
    title = "Missingness in liver proteome by sex and environment",
    x = NULL,
    y = "Fraction missing"
  )
```


```{r}
## Mean missingness by sex and environment.

miss_by_sample %>%
  group_by(sex, environment) %>%
  summarise(
    mean_missing = mean(frac_missing),
    sd_missing = sd(frac_missing),
    n = n(),
    .groups = "drop"
  )
```


```{r}
## Number of quantified proteins per sample after DEP filtering.

detected_per_sample <- colSums(!is.na(assay(data_filt)))

det_df <- data.frame(
  sample = names(detected_per_sample),
  detected = detected_per_sample
) %>%
  left_join(
    sample_meta %>%
      dplyr::select(sample, sex, age, environment),
    by = "sample"
  )

det_df %>%
  group_by(sex) %>%
  summarise(
    mean_detected = mean(detected),
    sd_detected = sd(detected),
    min_detected = min(detected),
    max_detected = max(detected),
    n = n(),
    .groups = "drop"
  )
```



```{r}
## Impute missing values using DEP2's MinProb method.

data_imp <- impute(
  data_norm,
  fun = "MinProb",
  q = 0.01
)

cat("Proteins after imputation:", nrow(data_imp), "\n")
cat("Samples after imputation:", ncol(data_imp), "\n")

cat(
  "Remaining missing values:",
  sum(is.na(assay(data_imp))),
  "\n"
)
```

```{r}
## Visualize intensity distributions after normalization and after imputation.

par(mfrow = c(1, 2))

boxplot(
  assay(data_norm),
  outline = FALSE,
  las = 2,
  main = "After VSN normalization",
  ylab = "Log2 intensity"
)

boxplot(
  assay(data_imp),
  outline = FALSE,
  las = 2,
  main = "After MinProb imputation",
  ylab = "Log2 intensity"
)

par(mfrow = c(1, 1))
```


```{r}
## Density distributions before and after imputation.

par(mfrow = c(1, 2))

plot(
  density(assay(data_norm)[, 1], na.rm = TRUE),
  main = "VSN normalized",
  xlab = "Log2 intensity"
)

for(i in 2:ncol(assay(data_norm))) {
  lines(
    density(assay(data_norm)[, i], na.rm = TRUE),
    col = rgb(0, 0, 0, 0.2)
  )
}

plot(
  density(assay(data_imp)[, 1]),
  main = "After MinProb imputation",
  xlab = "Log2 intensity"
)

for(i in 2:ncol(assay(data_imp))) {
  lines(
    density(assay(data_imp)[, i]),
    col = rgb(0, 0, 0, 0.2)
  )
}

par(mfrow = c(1, 1))
```

```{r}
## PCA using all filtered proteins after normalization and imputation.

mat_pca <- assay(data_imp)

pca_all <- prcomp(
  t(mat_pca),
  center = TRUE,
  scale. = TRUE
)

pca_var_all <- round(
  100 * (pca_all$sdev^2 / sum(pca_all$sdev^2)),
  1
)

pca_var_all[1:5]
```


```{r}
## PCA dataframe from imputed liver data.

pca_df <- as.data.frame(pca_all$x[, 1:2]) %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::bind_cols(
    as.data.frame(colData(data_imp)) %>%
      dplyr::select(-sample)
  )

head(
  pca_df[, c("sample_id", "sex", "age", "environment")]
)
```


```{r}
## Inspect PCA metadata columns using the available sample column name.

head(
  pca_df[, c("sample_id", "sex", "age", "environment")]
)
```


```{r}
## PCA coloured by sex and shaped by environment.

ggplot(
  pca_df,
  aes(
    PC1,
    PC2,
    color = sex,
    shape = environment
  )
) +
  geom_point(size = 4) +
  geom_text(
    aes(label = gsub("wk", "", age)),
    vjust = -0.8,
    size = 4
  ) +
  theme_bw() +
  labs(
    title = "liver PCA after filtering, VSN normalization and imputation",
    x = paste0("PC1 (", pca_var_all[1], "%)"),
    y = paste0("PC2 (", pca_var_all[2], "%)")
  )
```
 liver proteome reveals age-associated separation. PC1 explains 35.2% of the variance and clearly separates 8-week and 18-week samples, with mean PC1 scores of +27.0 and −33.8, respectively. This indicates that age may be a main driver. In contrast to the muscle dataset, sex and environment do not produce obvious global clustering patterns. PC2 explains 15.8% of the variance and is influenced by a subset of Mesocosm samples that exhibit distinct proteomic profiles, suggesting additional biological heterogeneity within the Mesocosm groups.



```{r}
## Mean PC2 score by age.

pca_df %>%
  dplyr::group_by(age) %>%
  dplyr::summarise(
    mean_PC2 = mean(PC2),
    sd_PC2 = sd(PC2),
    n = dplyr::n()
  )
```



```{r}
## Visualize overall protein abundance per sample by environment and age.

aprot_liver <- assay(data_imp) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Protein") %>%
  tidyr::pivot_longer(
    cols = -Protein,
    names_to = "SampleID",
    values_to = "PG_abundance"
  ) %>%
  dplyr::left_join(
    sample_meta %>%
      dplyr::select(
        SampleID = sample,
        age,
        environment,
        sex
      ),
    by = "SampleID"
  ) %>%
  dplyr::mutate(
    EnvAge = paste0(
      ifelse(environment == "Lab", "L", "M"),
      gsub("wk", "", age)
    ),
    EnvAge = factor(EnvAge, levels = c("L8", "M8", "L18", "M18")),
    AgeStage = factor(
      ifelse(age == "8wk", "Young", "Middle"),
      levels = c("Young", "Middle")
    ),
    Envt = factor(
      ifelse(environment == "Lab", "L", "M"),
      levels = c("L", "M")
    )
  )

ggplot(
  aprot_liver,
  aes(x = EnvAge, y = PG_abundance, color = Envt)
) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.15, size = 0.4) +
  facet_wrap(~ AgeStage, scales = "free_x", drop = TRUE) +
  theme_bw() +
  labs(
    title = "Overall liver protein abundance",
    x = "Environment and age",
    y = "Protein-group abundance",
    color = "Environment"
  )
```
Global protein abundance distributions are broadly similar across age and environment groups after normalization and imputation, indicating successful preprocessing and comparable overall proteome coverage. While median abundance is modestly higher in 18-week laboratory fish, differences in overall abundance distributions are small relative to the strong age-associated separation observed in PCA. These results suggest that aging primarily alters the composition of the liver proteome rather than causing a uniform increase or decrease in total protein abundance.


```{r}
## Test whether overall protein abundance differs between 8-week and 18-week liver samples.

subset_data <- aprot_liver %>%
  dplyr::filter(AgeStage %in% c("Young", "Middle"))

t_test_age <- t.test(
  PG_abundance ~ AgeStage,
  data = subset_data
)

t_test_age
```


```{r}
## Pairwise tests of overall protein abundance among L8, L8, L18, and L18.

pairwise.t.test(
  aprot_liver$PG_abundance,
  aprot_liver$EnvAge,
  p.adjust.method = "BH"
)
```
age signal is real and that environment effects may strengthen with age.

```{r}
## Volcano plot: M8 vs L8

group1 <- rownames(sample_meta)[sample_meta$condition == "L_8FL"]
group2 <- rownames(sample_meta)[sample_meta$condition == "L_8FM"]

mat <- assay(data_imp)

log2_fc <- rowMeans(mat[, group2]) - rowMeans(mat[, group1])

p_values <- apply(
  mat[, c(group1, group2)],
  1,
  function(x) {
    g1 <- x[1:length(group1)]
    g2 <- x[(length(group1) + 1):length(x)]
    t.test(g1, g2)$p.value
  }
)

v_data <- data.frame(
  Protein = rownames(mat),
  log2FC = log2_fc,
  p_value = p_values
)

v_data$neg_log10P <- -log10(v_data$p_value)

log2FC_thresh <- 1
pval_thresh <- -log10(0.05)

v_data$category <- with(
  v_data,
  ifelse(
    abs(log2FC) > log2FC_thresh & neg_log10P > pval_thresh,
    "Significant Both",
    ifelse(
      abs(log2FC) > log2FC_thresh,
      "Significant Log2FC",
      ifelse(
        neg_log10P > pval_thresh,
        "Significant P-value",
        "Not Significant"
      )
    )
  )
)

library(ggrepel)

ggplot(
  v_data,
  aes(log2FC, neg_log10P, color = category)
) +
  geom_point(alpha = 0.6) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  geom_text_repel(
    aes(
      label = ifelse(
        category == "Significant Both",
        Protein,
        ""
      )
    ),
    max.overlaps = 15,
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Not Significant" = "grey80",
      "Significant P-value" = "black",
      "Significant Log2FC" = "#009E73",
      "Significant Both" = "orange"
    )
  ) +
  labs(
    title = "Mesocosm vs Lab (8 wk)",
    x = "Log2 Fold Change",
    y = "-log10(p-value)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```


##improvements
```{r}
## Volcano plot: Mesocosm vs Lab at 8 weeks (female samples)
## Uses gene names and BH-adjusted p-values.

library(ggrepel)

group_lab <- rownames(sample_meta)[sample_meta$condition == "L_8FL"]
group_meso <- rownames(sample_meta)[sample_meta$condition == "L_8FM"]

mat <- assay(data_imp)

## Calculate log2 fold change.
log2_fc <- rowMeans(mat[, group_meso]) -
           rowMeans(mat[, group_lab])

## Protein-wise t-tests.
p_values <- apply(
  mat[, c(group_lab, group_meso)],
  1,
  function(x) {
    g1 <- x[1:length(group_lab)]
    g2 <- x[(length(group_lab) + 1):length(x)]
    t.test(g1, g2)$p.value
  }
)

## FDR correction.
padj <- p.adjust(p_values, method = "BH")

## Get gene names from DEP2 object.
gene_names <- rowData(data_imp)$name

v_data <- data.frame(
  Protein = rownames(mat),
  Gene = gene_names,
  log2FC = log2_fc,
  p_value = p_values,
  padj = padj
)

## Use protein ID if gene name missing.
v_data$Gene[is.na(v_data$Gene) | v_data$Gene == ""] <-
  v_data$Protein[is.na(v_data$Gene) | v_data$Gene == ""]

v_data$neg_log10_FDR <- -log10(v_data$padj)

## Thresholds.
fc_cutoff <- 1
fdr_cutoff <- 0.05

v_data$category <- with(
  v_data,
  ifelse(
    abs(log2FC) > fc_cutoff & padj < fdr_cutoff,
    "FDR + FC",
    ifelse(
      padj < fdr_cutoff,
      "FDR only",
      ifelse(
        abs(log2FC) > fc_cutoff,
        "FC only",
        "Not significant"
      )
    )
  )
)

## Label strongest FDR+FC hits.
label_df <- v_data %>%
  dplyr::filter(category == "FDR + FC") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 25)

n_sig <- sum(v_data$category == "FDR + FC")

ggplot(
  v_data,
  aes(log2FC, neg_log10_FDR, color = category)
) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_vline(
    xintercept = c(-fc_cutoff, fc_cutoff),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(fdr_cutoff),
    linetype = "dashed"
  ) +
  geom_text_repel(
    data = label_df,
    aes(label = Gene),
    size = 3,
    max.overlaps = 50
  ) +
  scale_color_manual(
    values = c(
      "Not significant" = "grey80",
      "FC only" = "#1b9e77",
      "FDR only" = "black",
      "FDR + FC" = "#d95f02"
    )
  ) +
  theme_bw() +
  labs(
    title = paste0(
      "Mesocosm vs Lab (8 wk females)\n",
      n_sig,
      " proteins pass FDR < 0.05 and |log2FC| > 1"
    ),
    x = "Log2 fold change (Mesocosm / Lab)",
    y = "-log10(FDR)"
  )
```




```{r}
## L18 vs L8 volcano using the DEP2-imputed liver dataset.

library(ggrepel)

group_M8 <- rownames(sample_meta)[sample_meta$condition == "L_8FM"]
group_M18 <- rownames(sample_meta)[sample_meta$condition == "L_18FM"]

mat <- assay(data_imp)

## Compute log2 fold changes.
log2_fc <- rowMeans(mat[, group_M18]) -
           rowMeans(mat[, group_M8])

## Protein-wise t-tests.
p_values <- apply(
  mat[, c(group_M8, group_M18)],
  1,
  function(x) {
    g1 <- x[1:length(group_M8)]
    g2 <- x[(length(group_M8) + 1):length(x)]
    t.test(g1, g2)$p.value
  }
)

## Gene names.
gene_names <- rowData(data_imp)$name

v_data <- data.frame(
  Protein = rownames(mat),
  Gene = gene_names,
  log2FC = log2_fc,
  p_value = p_values
)

v_data$Gene[is.na(v_data$Gene) | v_data$Gene == ""] <-
  v_data$Protein[is.na(v_data$Gene) | v_data$Gene == ""]

v_data$neg_log10P <- -log10(v_data$p_value)

## Thresholds.
log2FC_thresh <- 1
pval_thresh <- -log10(0.05)

v_data$category <- with(
  v_data,
  ifelse(
    abs(log2FC) > log2FC_thresh &
      neg_log10P > pval_thresh,
    "Significant Both",
    ifelse(
      abs(log2FC) > log2FC_thresh,
      "Significant Log2FC",
      ifelse(
        neg_log10P > pval_thresh,
        "Significant P-value",
        "Not Significant"
      )
    )
  )
)

## Volcano plot.
lm18 <- ggplot(
  v_data,
  aes(
    x = log2FC,
    y = neg_log10P,
    color = category
  )
) +
  geom_point(alpha = 0.6) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  geom_text_repel(
    aes(
      label = ifelse(
        category == "Significant Both",
        Gene,
        ""
      )
    ),
    box.padding = 0.5,
    max.overlaps = 20,
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Not Significant" = "gray85",
      "Significant P-value" = "gray5",
      "Significant Log2FC" = "#009E73",
      "Significant Both" = "orange"
    )
  ) +
  theme_bw() +
  labs(
    title = "Mesocosm 18 wk vs 8 wk",
    x = "Log2 Fold Change",
    y = "-log10(p-value)"
  ) +
  theme(
    legend.position = "none"
  )

lm18
```



```{r}
## Define a reusable function for volcano plots from the DEP2-imputed liver dataset.

library(ggrepel)

make_liver_volcano <- function(group1, group2, plot_title) {
  
  mat <- assay(data_imp)
  
  group1_samples <- rownames(sample_meta)[sample_meta$condition == group1]
  group2_samples <- rownames(sample_meta)[sample_meta$condition == group2]
  
  log2_fc <- rowMeans(mat[, group2_samples]) -
    rowMeans(mat[, group1_samples])
  
  p_values <- apply(
    mat[, c(group1_samples, group2_samples)],
    1,
    function(x) {
      g1 <- x[1:length(group1_samples)]
      g2 <- x[(length(group1_samples) + 1):length(x)]
      t.test(g1, g2)$p.value
    }
  )
  
  gene_names <- rowData(data_imp)$name
  
  v_data <- data.frame(
    Protein = rownames(mat),
    Gene = gene_names,
    log2FC = log2_fc,
    p_value = p_values
  )
  
  v_data$Gene[is.na(v_data$Gene) | v_data$Gene == ""] <-
    v_data$Protein[is.na(v_data$Gene) | v_data$Gene == ""]
  
  v_data$neg_log10P <- -log10(v_data$p_value)
  
  log2FC_thresh <- 1
  pval_thresh <- -log10(0.05)
  
  v_data$category <- with(
    v_data,
    ifelse(
      abs(log2FC) > log2FC_thresh & neg_log10P > pval_thresh,
      "Significant Both",
      ifelse(
        abs(log2FC) > log2FC_thresh,
        "Significant Log2FC",
        ifelse(
          neg_log10P > pval_thresh,
          "Significant P-value",
          "Not Significant"
        )
      )
    )
  )
  
  label_df <- v_data %>%
    dplyr::filter(category == "Significant Both") %>%
    dplyr::arrange(p_value) %>%
    dplyr::slice_head(n = 20)
  
  p <- ggplot(
    v_data,
    aes(x = log2FC, y = neg_log10P, color = category)
  ) +
    geom_point(alpha = 0.6) +
    geom_vline(
      xintercept = c(-log2FC_thresh, log2FC_thresh),
      linetype = "dashed"
    ) +
    geom_hline(
      yintercept = pval_thresh,
      linetype = "dashed"
    ) +
    geom_text_repel(
      data = label_df,
      aes(label = Gene),
      box.padding = 0.5,
      max.overlaps = 20,
      size = 3
    ) +
    scale_color_manual(
      values = c(
        "Not Significant" = "gray85",
        "Significant P-value" = "gray5",
        "Significant Log2FC" = "#009E73",
        "Significant Both" = "orange"
      )
    ) +
    theme_bw() +
    labs(
      title = plot_title,
      x = paste0("Log2 fold change (", group2, " / ", group1, ")"),
      y = "-log10(p-value)"
    ) +
    theme(
      legend.position = "none"
    )
  
  return(
    list(
      plot = p,
      data = v_data
    )
  )
}
```


```{r}
## Generate the four main liver volcano plots.

volc_M8_vs_L8 <- make_liver_volcano(
  group1 = "L_8FL",
  group2 = "L_8FM",
  plot_title = "Mesocosm vs Lab at 8 wk"
)

volc_M18_vs_L18 <- make_liver_volcano(
  group1 = "L_18FL",
  group2 = "L_18FM",
  plot_title = "Mesocosm vs Lab at 18 wk"
)

volc_L18_vs_L8 <- make_liver_volcano(
  group1 = "L_8FL",
  group2 = "L_18FL",
  plot_title = "Lab aging: 18 wk vs 8 wk"
)

volc_M18_vs_M8 <- make_liver_volcano(
  group1 = "L_8FM",
  group2 = "L_18FM",
  plot_title = "Mesocosm aging: 18 wk vs 8 wk"
)
```





```{r}
## Display the Mesocosm aging volcano plot.

volc_M18_vs_M8$plot
```


```{r}
## Arrange the four volcano plots into a 2 x 2 panel.

library(ggpubr)

volcano_panel <- ggarrange(
  volc_M8_vs_L8$plot,
  volc_M18_vs_L18$plot,
  volc_L18_vs_L8$plot,
  volc_M18_vs_M8$plot,
  nrow = 2,
  ncol = 2,
  labels = c("A", "B", "C", "D")
)

volcano_panel
```
Liver aging appears to be the dominant proteomic process in this dataset, producing substantially larger proteomic shifts than environmental differences, whereas environmental effects are already detectable at 8 weeks and persist, albeit more weakly, at 18 weeks.


```{r}
nrow(
  subset(
    volc_M8_vs_L8$data,
    abs(log2FC) > 1 & p_value < 0.05
  )
)

nrow(
  subset(
    volc_M18_vs_L18$data,
    abs(log2FC) > 1 & p_value < 0.05
  )
)

nrow(
  subset(
    volc_L18_vs_L8$data,
    abs(log2FC) > 1 & p_value < 0.05
  )
)

nrow(
  subset(
    volc_M18_vs_M8$data,
    abs(log2FC) > 1 & p_value < 0.05
  )
)
```


```{r}
## Build liver limma model fresh.

library(limma)

sample_meta <- as.data.frame(colData(data_imp))

sample_meta$condition <- factor(sample_meta$condition)

design <- model.matrix(
  ~ 0 + condition,
  data = sample_meta
)

colnames(design) <- levels(sample_meta$condition)

cat("Design dimensions:\n")
print(dim(design))

cat("Design columns:\n")
print(colnames(design))

fit <- limma::lmFit(
  assay(data_imp),
  design
)


cont.matrix
cont.matrix <- limma::makeContrasts(
  M8_vs_L8   = L_8FM  - L_8FL,
  M18_vs_L18 = L_18FM - L_18FL,
  L18_vs_L8  = L_18FL - L_8FL,
  M18_vs_M8  = L_18FM - L_8FM,
  Age_by_Environment =
    (L_18FM - L_8FM) -
    (L_18FL - L_8FL),
  levels = design
)

fit2 <- limma::contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

cat("Fit2 dimensions:\n")
print(dim(fit2$coefficients))

summary(decideTests(fit2))
```

Aging is the dominant driver of liver proteomic remodeling, with especially strong remodeling in Lab females. Environment alone has little FDR-significant effect, and the age-by-environment interaction does not reach significance.


```{r}
## Top proteins changing with Mesocosm aging.

top_M18_vs_M8 <- topTable(
  fit2,
  coef = "M18_vs_M8",
  number = 20,
  sort.by = "P"
)

top_M18_vs_M8
```



```{r}
## Examine significant proteins.

sig_M18_vs_M8 <- topTable(
  fit2,
   coef = "M18_vs_M8",
   number = Inf,
    p.value = 0.05)

nrow(sig_M18_vs_M8) 
```


```{r}
lab_sig <- rownames(
  topTable(
    fit2,
    coef = "L18_vs_L8",
    number = Inf,
    p.value = 0.05
  )
)

meso_sig <- rownames(
  topTable(
    fit2,
    coef = "M18_vs_M8",
    number = Inf,
    p.value = 0.05
  )
)

shared_sig <- intersect(lab_sig, meso_sig)

length(shared_sig)
```



```{r}
lab_only <- setdiff(lab_sig, meso_sig)

meso_only <- setdiff(meso_sig, lab_sig)

length(lab_only)
length(meso_only)
```


```{r}
 shared_sig <- intersect(lab_sig, meso_sig)

lab_only <- setdiff(lab_sig, meso_sig)

meso_only <- setdiff(meso_sig, lab_sig)
```



```{r}
shared_clean <- unique(clean_gene_symbol(shared_sig))
lab_only_clean <- unique(clean_gene_symbol(lab_only))
meso_only_clean <- unique(clean_gene_symbol(meso_only))
```


```{r}
length(shared_sig)
length(shared_clean)

length(lab_only)
length(lab_only_clean)

length(meso_only)
length(meso_only_clean)

head(shared_clean, 20)
```


```{r}
shared_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = shared_clean,
  keytype = "SYMBOL",
  columns = c("SYMBOL", "ENTREZID")
)
```


```{r}
shared_map_clean <- shared_map %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::distinct(SYMBOL, ENTREZID)

cat("Shared cleaned genes:", length(shared_clean), "\n")
cat("Mapped genes:", nrow(shared_map_clean), "\n")
cat(
  "Mapping percentage:",
  round(100 * nrow(shared_map_clean) / length(shared_clean), 1),
  "%\n"
)

head(shared_map_clean)
```

##mapping rate
```{r}
cat("Shared cleaned genes:", length(shared_clean), "\n")
cat("Mapped genes:", nrow(shared_map_clean), "\n")
cat(
  "Mapping percentage:",
  round(100 * nrow(shared_map_clean) / length(shared_clean), 1),
  "%\n"
)
```


```{r}
ego_shared <- enrichGO(
  gene          = shared_map_clean$ENTREZID,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)
```


```{r}
nrow(as.data.frame(ego_shared))
head(
  as.data.frame(ego_shared)[,
    c("Description", "GeneRatio", "p.adjust", "Count")
  ],
  20
)
```
Aging in the liver induces broad metabolic remodeling involving lipid utilization, amino acid metabolism, mitochondrial energy production, and protein synthesis.


##visualize

```{r}
dotplot(
  ego_shared,
  showCategory = 20,
  font.size = 12
)
```


```{r}
ego_shared_simp <- simplify(
  ego_shared,
  cutoff = 0.7,
  by = "p.adjust",
  select_fun = min
)

dotplot(
  ego_shared_simp,
  showCategory = 20,
  font.size = 12
)
```

To identify biological processes associated with liver aging independent of environmental conditions, I examined the 868 proteins that were significantly altered with age in both laboratory and mesocosm fish. After collapsing protein identifiers and mapping to human orthologs, 440 genes (66%) were successfully annotated and used for GO enrichment analysis. Shared aging proteins were strongly enriched for metabolic processes, including small molecule catabolism (FDR = 1.8 × 10⁻²²), organic acid catabolism (FDR = 7.0 × 10⁻²¹), amino acid metabolism (FDR = 2.1 × 10⁻¹¹), and fatty acid oxidation (FDR = 3.1 × 10⁻¹¹). Pathways related to cellular respiration and energy generation were also enriched, indicating extensive remodeling of hepatic energy metabolism with age. In addition, terms associated with cytoplasmic translation, protein folding, endoplasmic reticulum stress responses, and regulation of protein stability were enriched, suggesting age-related changes in proteostasis and secretory pathway function. These results indicate that liver aging is characterized by coordinated metabolic and proteostatic remodeling that is largely conserved across environmental conditions.




```{r}
## Extract the top 50 significant proteins from Mesocosm aging.

top50_ids <- rownames(
  topTable(
    fit2,
    coef = "M18_vs_M8",
    number = 50,
    sort.by = "P"
  )
)

length(top50_ids)

head(top50_ids)
```

```{r}
## Heatmap of the top 100 shared liver aging proteins.
## These proteins are significantly altered with age in both Lab and Mesocosm fish.

library(pheatmap)

shared_top100 <- shared_sig[
  order(
    abs(fit2$coefficients[shared_sig, "L18_vs_L8"]),
    decreasing = TRUE
  )
][1:50]

heat_mat <- assay(data_imp)[shared_top100, ]

heat_mat_z <- t(
  scale(
    t(heat_mat)
  )
)

annotation_col <- as.data.frame(colData(data_imp))[
  , c("age", "environment", "sex")
]

pheatmap(
  heat_mat_z,
  annotation_col = annotation_col,
  show_rownames = FALSE,
  fontsize_col = 10,
  clustering_method = "complete",
  main = "Shared liver aging proteins (top 100)"
)
```







```{r}
## Heatmap of the top 50 Mesocosm-aging proteins.

library(pheatmap)

meso_samples <- rownames(sample_meta)[
  sample_meta$condition %in% c("L_8FM", "L_18FM")
]

heat_mat <- assay(data_imp)[top50_ids, meso_samples]

## Z-score by protein.
heat_mat_z <- t(scale(t(heat_mat)))

annotation_col <- data.frame(
  Age = sample_meta[meso_samples, "age"],
  row.names = meso_samples
)

pheatmap(
  heat_mat_z,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  show_colnames = TRUE,
  clustering_method = "complete",
  fontsize_row = 6,
  main = "Top 50 proteins: Mesocosm aging (18 wk vs 8 wk)"
)
```

Hierarchical clustering of the top 50 proteins differentially abundant between 8- and 18-week Mesocosm females reveal separation of age groups. Older fish exhibit increased abundance of structural and contractile proteins, including collagens and myofibrillar proteins, whereas numerous ribosomal and ribosome-biogenesis-associated proteins are reduced.

#heatmap all 181
```{r}
## Heatmap of all significant proteins from Mesocosm aging.

library(pheatmap)

sig_ids <- rownames(sig_M18_vs_M8)

meso_samples <- rownames(sample_meta)[
  sample_meta$condition %in% c("L_8FM", "L_18FM")
]

heat_mat <- assay(data_imp)[sig_ids, meso_samples]

## Z-score by protein.
heat_mat_z <- t(scale(t(heat_mat)))

annotation_col <- data.frame(
  Age = sample_meta[meso_samples, "age"],
  row.names = meso_samples
)

pheatmap(
  heat_mat_z,
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_method = "complete",
  fontsize_col = 10,
  main = paste0(
    "Mesocosm aging signature (",
    length(sig_ids),
    " proteins)"
  )
)
```

separates into M_8FM samples vs M_18FM samples, no mixing
Differential abundance analysis identified 181 proteins significantly altered between 8- and 18-week mesocosm females (FDR < 0.05). Hierarchical clustering based on these proteins completely separated age groups, indicating age-associated proteomic signature. Proteins elevated in younger fish were enriched for ribosomal and translation-associated functions, whereas proteins elevated in older fish included contractile and extracellular matrix components.

## Split heatmaps for proteins increased and decreased with Mesocosm aging.
```{r}


library(pheatmap)

meso_samples <- rownames(sample_meta)[
  sample_meta$condition %in% c("L_8FM", "L_18FM")
]

annotation_col <- data.frame(
  Age = sample_meta[meso_samples, "age"],
  row.names = meso_samples
)

## Up with age
up_ids <- rownames(up_M18_vs_M8)

heat_up <- assay(data_imp)[up_ids, meso_samples]
heat_up_z <- t(scale(t(heat_up)))

pheatmap(
  heat_up_z,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  show_colnames = TRUE,
  clustering_method = "complete",
  fontsize_row = 7,
  main = paste0(
    "Proteins increased with Mesocosm aging (",
    length(up_ids),
    " proteins)"
  )
)

## Down with age
down_ids <- rownames(down_M18_vs_M8)

heat_down <- assay(data_imp)[down_ids, meso_samples]
heat_down_z <- t(scale(t(heat_down)))

pheatmap(
  heat_down_z,
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_method = "complete",
  fontsize_col = 10,
  main = paste0(
    "Proteins decreased with Mesocosm aging (",
    length(down_ids),
    " proteins)"
  )
)
```
aging signal is much stronger than within-group variation, there is a coordinated reduction in the protein synthesis machinery during mesocosm liver aging
Higher abundance at 8 wk
Lower abundance at 18 wk

##important question still to ask (M18−M8)−(L18−L8): Which proteins change with age in Mesocosm but not in Lab?/Which proteins age differently between environments?
## so frar results hint that there is little age-related proteomic remodeling in lab and there are strong age-related remodeling in meso
#add contrasts
```{r}
cont.matrix <- makeContrasts(
  M8_vs_L8   = L_8FM  - L_8FL,
  M18_vs_L18 = L_18FM - L_18FL,
  L18_vs_L8  = L_18FL - L_8FL,
  M18_vs_M8  = L_18FM - L_8FM,

  Age_by_Environment =
    (L_18FM - L_8FM) -
    (L_18FL L_8FL),

  levels = design
)
```

#Refit
```{r}
up_M18_vs_M8 <- sig_M18_vs_M8[
  sig_M18_vs_M8$logFC > 0,
]

down_M18_vs_M8 <- sig_M18_vs_M8[
  sig_M18_vs_M8$logFC < 0,
]

nrow(up_M18_vs_M8)
nrow(down_M18_vs_M8)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

summary(decideTests(fit2))
```
        


```{r}
top_interaction <- topTable(
  fit2,
  coef = "Age_by_Environment",
  number = 20
)

top_interaction[, c("logFC", "P.Value", "adj.P.Val")]
```

#visualise the genes that show near-significant interaction effects


```{r}

## Visualize multiple interaction-candidate proteins across female groups.
## Age is grouped first, with Lab and Mesocosm shown within each age.

genes_to_plot <- c(
)

plot_multi_df <- assay(data_imp)[genes_to_plot, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Protein") %>%
  tidyr::pivot_longer(
    cols = -Protein,
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  dplyr::left_join(
    sample_meta %>%
      dplyr::select(sample, condition, age, environment, sex),
    by = "sample"
  ) %>%
  dplyr::filter(
    sex == "Female",
    condition %in% c("L_8FL", "L_18FL", "L_8FM", "L_18FM")
  ) %>%
  dplyr::mutate(
    age = factor(age, levels = c("8wk", "18wk")),
    environment = factor(environment, levels = c("Lab", "Mesocosm"))
  )

ggplot(
  plot_multi_df,
  aes(x = age, y = abundance, fill = environment)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.6,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    aes(color = environment),
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 1.8,
    alpha = 0.85
  ) +
  facet_wrap(~ Protein, scales = "free_y", ncol = 3) +
  theme_bw() +
  labs(
    title = "Candidate age-by-environment proteins in female liver",
    x = "Age",
    y = "Normalized/imputed abundance",
    fill = "Environment",
    color = "Environment"
  ) +
  theme(
    legend.position = "bottom"
  )
  
```



#####summary???
Aging-associated proteomic remodeling is most pronounced in mesocosm females. Differential abundance analysis identifies 181 proteins that change significantly between 8- and 18-week mesocosm females (FDR < 0.05), whereas no proteins reach significance in the corresponding laboratory aging comparison. Proteins that decrease with age are strongly enriched for cytoplasmic translation, ribosome biogenesis, and rRNA processing pathways, while proteins that increase with age are enriched for liver contraction and liver system processes.

To determine whether aging trajectories differ significantly between environments, I tested an age-by-environment interaction contrast: (M_18FM − M_8FM) − (M_18FL − M_8FL). No proteins reach the FDR < 0.05 threshold in this formal interaction analysis. However, several biologically relevant proteins show near-significant interaction effects, including PVALB7 (interaction logFC = 4.18, FDR = 0.052), Nfu_g_1_015914.1 (logFC = 2.83, FDR = 0.052), MYLPFA (logFC = 4.47, FDR = 0.063), TPP2.1 (logFC = -3.04, FDR = 0.067), and Nfu_g_1_009250 (logFC = 3.01, FDR = 0.073). Additional proteins associated with liver structure and contraction, including CASQ1A (FDR = 0.104), COL1A1B (FDR = 0.146), and MYH7B (FDR = 0.146), also rank among the strongest interaction candidates.

Together, these results indicate age-associated proteomic remodeling in mesocosm liver, characterized by reduced abundance of translation- and ribosome-associated proteins and increased abundance of contractile and structural liver proteins. Although age-by-environment interactions do not survive multiple-testing correction, the near-significant interaction effects suggest that environmental context influences the magnitude of age-related proteomic remodeling and warrant further investigation in larger cohorts, may be???

##plot those with strongest interaction
```{r}
## Strongest candidate age-by-environment proteins.

genes_to_plot <- c(
  
)

plot_df <- assay(data_imp)[genes_to_plot, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Protein") %>%
  tidyr::pivot_longer(
    cols = -Protein,
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  dplyr::left_join(
    sample_meta %>%
      dplyr::select(
        sample,
        condition,
        age,
        environment,
        sex
      ),
    by = "sample"
  ) %>%
  dplyr::filter(
    sex == "Female",
    condition %in% c(
      "L_8FL",
      "L_18FL",
      "L_8FM",
      "L_18FM"
    )
  ) %>%
  dplyr::mutate(
    age = factor(age, levels = c("8wk", "18wk")),
    environment = factor(
      environment,
      levels = c("Lab", "Mesocosm")
    )
  )

ggplot(
  plot_df,
  aes(
    x = age,
    y = abundance,
    fill = environment
  )) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.6,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    aes(color = environment),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.75
    ),
    size = 2,
    alpha = 0.85
  ) +
  facet_wrap(
    ~ Protein,
    scales = "free_y",
    ncol = 2
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Top candidate age-by-environment proteins",
    x = "Age",
    y = "Normalized/imputed abundance",
    fill = "Environment",
    color = "Environment"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```
PVALB7, MYLPFA, COL1A1B and MRPL39.1 were selected from the age-by-environment interaction analysis as representative candidate proteins. Interaction FDR values were 0.052, 0.063, 0.146 and 0.103, respectively.


```{r}
#strongest candidate age-by-environment proteins.
## Shows age comparisons within Lab and Mesocosm females,
## with interaction FDRs included in facet labels.

library(ggpubr)
library(rstatix)

genes_to_plot <- c(
  
)

plot_df <- assay(data_imp)[genes_to_plot, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Protein") %>%
  tidyr::pivot_longer(
    cols = -Protein,
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  dplyr::left_join(
    sample_meta %>%
      dplyr::select(sample, condition, age, environment, sex),
    by = "sample"
  ) %>%
  dplyr::filter(
    sex == "Female",
    condition %in% c("L_8FL", "L_18FL", "L_8FM", "L_18FM")
  ) %>%
  dplyr::mutate(
    age = factor(age, levels = c("8wk", "18wk")),
    environment = factor(environment, levels = c("Lab", "Mesocosm")),
    group = factor(
      paste(environment, age),
      levels = c(
        "Lab 8wk",
        "Lab 18wk",
        "Mesocosm 8wk",
        "Mesocosm 18wk"
      )
    ),
    Protein = factor(
      Protein,
      levels = c(),
      labels = c(
        
      )
    )
  )

stat_df <- plot_df %>%
  dplyr::group_by(Protein, environment) %>%
  rstatix::t_test(abundance ~ age) %>%
  rstatix::add_significance("p") %>%
  dplyr::mutate(
    group1 = ifelse(environment == "Lab", "Lab 8wk", "Mesocosm 8wk"),
    group2 = ifelse(environment == "Lab", "Lab 18wk", "Mesocosm 18wk")
  ) %>%
  dplyr::group_by(Protein) %>%
  dplyr::mutate(
    y.position = max(plot_df$abundance[plot_df$Protein == Protein[1]]) +
      dplyr::row_number() * 0.45
  ) %>%
  dplyr::ungroup()

candidate_plot <- ggplot(
  plot_df,
  aes(x = group, y = abundance, fill = environment)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.6
  ) +
  geom_jitter(
    aes(color = environment),
    width = 0.12,
    size = 2,
    alpha = 0.85
  ) +
  facet_wrap(
    ~ Protein,
    scales = "free_y",
    ncol = 2
  ) +
  
  stat_pvalue_manual(
  stat_df,
  label = "p.signif",
  xmin = "group1",
  xmax = "group2",
  y.position = "y.position",
  tip.length = 0.01
)
   +
  scale_x_discrete(
    labels = c(
      "Lab\n8wk",
      "Lab\n18wk",
      "Mesocosm\n8wk",
      "Mesocosm\n18wk"
    )
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Proteins exhibiting stronger age-related changes in Mesocosm females",
    x = NULL,
    y = "Normalized/imputed abundance",
    fill = "Environment",
    color = "Environment"
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold")
  )

candidate_plot
   

```

"PVALB7 (FDR = 0.052) and MYLPFA (FDR = 0.063) exhibit evidence for age-by-environment interactions. MRPL39.1 (FDR = 0.103) and COL1A1B (FDR = 0.146) show similar qualitative patterns but provide weaker statistical support."

###look @ others
```{r}
sum(top_interaction$adj.P.Val < 0.05)
sum(top_interaction$adj.P.Val < 0.10)
sum(top_interaction$adj.P.Val < 0.15)
sum(top_interaction$adj.P.Val < 0.20)
```



```{r}
## Extract interaction candidate genes.

interaction_genes <- unique(
  rowData(data_imp)[rownames(interaction20), "name"]
)

interaction_genes <- interaction_genes[
  !is.na(interaction_genes) &
  interaction_genes != ""
]

sort(interaction_genes)
length(interaction_genes)
```


```{r}
clean_gene_symbol <- function(x) {
  x <- toupper(x)
  x <- sub("\\.[0-9]+$", "", x)
  x
}

interaction_genes_clean <- unique(
  clean_gene_symbol(interaction_genes)
)

length(interaction_genes_clean)

interaction_genes_clean
```


```{r}
interaction_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = interaction_genes_clean,
  keytype = "SYMBOL",
  columns = c("SYMBOL", "ENTREZID")
)

interaction_map <- interaction_map %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::distinct(SYMBOL, ENTREZID)

cat(
  "Mapped:",
  nrow(interaction_map),
  "of",
  length(interaction_genes_clean),
  "genes\n"
)

interaction_map
```


```{r}
nrow(interaction_map)
head(interaction_map)
```



#save files
```{r}
## Significant proteins increased with age.

up_M18_vs_M8 <- sig_M18_vs_M8 %>%
  dplyr::filter(logFC > 0)

## Significant proteins decreased with age.

down_M18_vs_M8 <- sig_M18_vs_M8 %>%
  dplyr::filter(logFC < 0)

cat("Up with age:", nrow(up_M18_vs_M8), "\n")
cat("Down with age:", nrow(down_M18_vs_M8), "\n")

head(rownames(up_M18_vs_M8))
head(rownames(down_M18_vs_M8))
```

#genes with recognizable symbols
```{r}
sum(
  !grepl(
    "^Nfu_g_|^LOC|^G4P62_|^A0A",
    rownames(sig_M18_vs_M8)
  )
)
```

###use actual gene names stored in dep2 rather than rownames 
```{r}
## Get gene annotations for significant proteins.

sig_genes <- rowData(data_imp)[rownames(sig_M18_vs_M8), "name"]

head(sig_genes)
```


```{r}
## Up-regulated genes.

up_genes <- unique(
  rowData(data_imp)[rownames(up_M18_vs_M8), "name"]
)

up_genes <- up_genes[
  !is.na(up_genes) &
  up_genes != ""
]

length(up_genes)

## Down-regulated genes.

down_genes <- unique(
  rowData(data_imp)[rownames(down_M18_vs_M8), "name"]
)

down_genes <- down_genes[
  !is.na(down_genes) &
  down_genes != ""
]

length(down_genes)
```


#many unique gene symbols in each direction
```{r}
## Up-regulated genes.

up_genes <- unique(
  rowData(data_imp)[rownames(up_M18_vs_M8), "name"]
)

up_genes <- up_genes[
  !is.na(up_genes) &
  up_genes != ""
]

length(up_genes)

## Down-regulated genes.

down_genes <- unique(
  rowData(data_imp)[rownames(down_M18_vs_M8), "name"]
)

down_genes <- down_genes[
  !is.na(down_genes) &
  down_genes != ""
]

length(down_genes)
```
Identical numbers because there are no duplicated gene names among the significant proteins.

#quantify the ribosomal signal across all 181 proteins
```{r}
sum(
  grepl(
    "^RPL|^RPS|^MRPL|^MRPS",
    rownames(sig_M18_vs_M8)
  )
)
```

#what fraction of the aging signature they represent
```{r}
round(
  100 *
    sum(
      grepl(
        "^RPL|^RPS|^MRPL|^MRPS",
        rownames(sig_M18_vs_M8)
      )
    ) /
    nrow(sig_M18_vs_M8),
  1
)
```

Mesocosm aging associated with significant remodeling of the translational machinery. Among the 181 proteins differentially abundant between 8- and 18-week Mesocosm females (FDR < 0.05), 43 (23.8%) were annotated as cytosolic or mitochondrial ribosomal proteins. Most ribosomal proteins decreased in abundance with age, suggesting reduced translational capacity and/or altered ribosome biogenesis in older liver.


#enrichment analysis
```{r}
#clean up
## Examine the significant gene names.

head(sort(unique(all_sig_genes)), 50)

## Count genes with isoform suffixes.

sum(grepl("\\.[0-9]+$", all_sig_genes))
```

###human-symbol-based enrichment of killifish gene names- zebra fish mapping only a few
```{r}
## Map cleaned significant liver genes to human Entrez IDs.

human_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = all_sig_genes_clean,
  keytype = "SYMBOL",
  columns = c("SYMBOL", "ENTREZID")
)

human_map_clean <- human_map %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::distinct(SYMBOL, ENTREZID)

cat("Cleaned input genes:", length(all_sig_genes_clean), "\n")
cat("Human mapped genes:", nrow(human_map_clean), "\n")
cat(
  "Human mapping percentage:",
  round(100 * nrow(human_map_clean) / length(all_sig_genes_clean), 1),
  "%\n"
)

head(human_map_clean)
```


```{r}
## Prepare cleaned up- and down-regulated gene lists for human enrichment.

up_genes_clean <- unique(clean_gene_symbol(up_genes))
down_genes_clean <- unique(clean_gene_symbol(down_genes))

up_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = up_genes_clean,
  keytype = "SYMBOL",
  columns = c("SYMBOL", "ENTREZID")
) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::distinct(SYMBOL, ENTREZID)

down_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = down_genes_clean,
  keytype = "SYMBOL",
  columns = c("SYMBOL", "ENTREZID")
) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::distinct(SYMBOL, ENTREZID)

cat("Up genes:", length(up_genes_clean), "mapped:", nrow(up_map), "\n")
cat("Down genes:", length(down_genes_clean), "mapped:", nrow(down_map), "\n")
```
Up:   47 genes → 18 mapped (interpret this enrichment with care.)
Down: 128 genes → 96 mapped


```{r}
## GO Biological Process enrichment for proteins decreased with age in Mesocosm.

ego_down <- enrichGO(
  gene          = down_map$ENTREZID,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)

ego_down_df <- as.data.frame(ego_down)

head(
  ego_down_df[, c("Description", "GeneRatio", "p.adjust", "Count")],
  20
)
```

These terms are kind of suprising: myoblast fusion,myotube differentiation, striated liver contraction, liver cell differentiation and liver contraction
Because they came from the down-regulated list. Looks contradictory because saw proteins like:MYH13,MYLPFA,COL12A1 and PVALB7 increase with age!

May be many ribosomal proteins are annotated to liver-development pathways because those pathways contain large numbers of translation-related genes??

General conclusion: Proteins that decreased with age in Mesocosm females are strongly enriched for ribosomal proteins, cytoplasmic translation, ribosome biogenesis, and rRNA processing pathways.

##enrich the up-regulated proteins
```{r}
ego_up <- enrichGO(
  gene          = up_map$ENTREZID,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.20,
  readable      = TRUE
)

ego_up_df <- as.data.frame(ego_up)

head(
  ego_up_df[, c("Description", "GeneRatio", "p.adjust", "Count")],
  20
)
```
cytoplasmic translation, 43/96 genes, FDR ≈ 4.7 × 10^-55
Dominant biological process decreasing with age in Mesocosm liver!
liver contraction
liver system process
regulation of liver contraction





```{r}
## Dotplot of GO enrichment for proteins decreased with age.

library(enrichplot)

dotplot(
  ego_down,
  showCategory = 15,
  font.size = 12
)
```
Redundant terms:negative regulation of myoblast fusion
regulation of myoblast fusion
myoblast fusion
cell-cell fusion
syncytium formation

#Collapse redundant terms
```{r}
ego_down_simp <- simplify(
  ego_down,
  cutoff = 0.7,
  by = "p.adjust",
  select_fun = min
)

dotplot(
  ego_down_simp,
  showCategory = 15,
  font.size = 12
)
```


```{r}
library(enrichplot)

emapplot(
  pairwise_termsim(ego_down_simp),
  showCategory = 20
)
```


```{r}
## Dotplot of GO enrichment for proteins increased with age.

dotplot(
  ego_up,
  showCategory = 10,
  font.size = 12
)
```

Generally, mesocosm liver aging is characterized by a broad reduction in translational capacity accompanied by increased abundance of contractile and liver-structure proteins.


```{r}
write.csv(sig_M18_vs_M8,
          "Mesocosm_Aging_181_Proteins.csv")

write.csv(as.data.frame(ego_down),
          "Mesocosm_Aging_GO_Down.csv")

write.csv(as.data.frame(ego_up),
          "Mesocosm_Aging_GO_Up.csv")
```



```{r}

```


```{r}

```


```{r}

```


```{r}

```



```{r}

```

```{r}

```


```{r}

```


```{r}

```


